---
title: "Solar mini-grid power consumption profiles in DRC"
author:
  - name: "Elyse McMillen"
    orcid: "0009-0002-2827-3062"
    email: "elyse.mcmillen@colorado.edu"
    affiliation:
      - name: "University of Colorado Boulder"
        department: "Department of Civil, Environmental and Architectural Engineering"
        city: "Boulder"
        state: "CO"
        country: "USA"
date: "July 02 2026"
format: html
editor: visual

---

# Introduction

This project explores existing power consumption data over a set range of dates from April - May 2026. The data was gathered from an operational solar mini-grid in the DRC providing power to a local community. The goal of the project is to summarize the consumption by category to be able to scale the expected consumption for other sites to predict mini-grid sizing in conflict-ridden areas with minimal on-the-ground data collection. The intent is to group by category and type for the client and be able to summarize statistics on daily and peak power consumption to feed forward to other grid planning projects.

# Methods

The data was collected from the IT team at the organization after specifying the intent of the data. After receiving the initial data from the IT team that was gathered from smart meters, the raw data was examined and processed for anonymity. No financial information was included in the pulls. The names of the clients were initially included so that type of enterprise could be inferred in some cases (i.e. school or hospital) then the names were removed before adding the raw data. The data needs to be sorted, matched with business types, and collated to gather usage statistics. 

## Reading the Data
```{r}
#| message: false
#| warning: false
# Install packages
install.packages("readxl")
install.packages("writexl")
install.packages("polyglotr")
```

Since the file is a .xlsx file and not a .csv file, I had to install the "readxl" and "writexl" packages. I had some initial challenges loading the "readxl" package but finally was able to get it to work after uninstalling and re-installing. In hindsight, converting the file to a .csv might have been an easier way to start. 

```{r}
#| message: false
#| warning: false
# Load packages
library(tidyverse)
library(readxl)
library(dplyr)
library (writexl)
library (polyglotr)
library(knitr)
library(ggplot2)

# Read data
raw_power_data <- read_excel(here::here("data/raw/Power_consumption_categorized_EN_April_May_2026.xlsx"))

```

## Data Exploration Approach

Time to view the data and see what needs tidying!

```{r}
glimpse(raw_power_data)

dim(raw_power_data)

# calculate average daily kWh per customer type
customer_grouped <- raw_power_data %>% 
  group_by(`Customer type`) %>% 
  summarize(mean_daily_kWh = mean(`Daily kWh`, total_count = n()))

# 4 main types of customers to examine

# view the existing customer categories
category_grouped <- raw_power_data %>% 
  group_by(Category) %>% 
  summarize(mean_daily_kWh = mean(`Daily kWh`, total_count = n()))

# 21 groupings of client type with drastically different power consumption

```

The initial data set is 165,588 rows and 7 columns. 
The column names are "date", "category", "customer code", "customer type", "hourly kWh average", "hourly kWh Max", and "daily kWh". I see some initial values showing of 0.000000 under some entries, and many repeating categories or types. 

Examined the average daily kWh for each major category and type for reference.

## Initial Data Tidying

Fortunately the dates are already in YYYY-MM-DD format, so R has been able to ingest those ok. 

The data needs to sort out any 0.000000 and NA entries to not skew the mean values for the categories. The column names that have spaces need to be renames to remove the need to include "`" in every call-out.

```{r}

clean_power_data <- raw_power_data %>% 
  rename(
    customer_type = `Customer type`,
    category = Category,
    customer_code = `Customer Code`,
    customer_type = `Customer type`,
    kWh_avg_hourly = `Hourly kWh Average`,
    kWh_avg_max = `Hourly kWh Max`,
    kWh_per_day = `Daily kWh`
  )

# convert NA text to NA data entry in the sheet
power_na_clean <- clean_power_data %>% 
  mutate(across(where(is.character), ~ na_if(., "NA")))

# count the number of NA data entries
na_counts <- power_na_clean %>%
  group_by(customer_type) %>%
  summarise(na_count = sum(is.na(category)))

# Remove rows with all 0.0000 values (some 0 values are OK since it's by day
# , not every client will use power every day)
power_filtered <- power_na_clean %>%
  filter(if_any(c(kWh_avg_hourly, kWh_avg_max, kWh_per_day), ~ .x != 0))

```

There are >32,000 NA category entries in SME type. Removing all of these entries will skew the data significantly. Instead, keep NA category and treat as a general category.

The first time I tried to calculate the NA values I got 0... which is incorrect because I noted a bunch of NA values in the Category column. I first had to convert the text "NA" entries into actual NA entries.

Sorting out the entries with all 0 valued for the three numerical columns, the data set shrank from 165,588 rows to 149,263 rows. This now just includes data that has usable numerical values.


```{r}
# save as a .xlsx file 
write_xlsx(power_filtered, here::here("data/processed/power_filtered.xlsx"))

```

# Results

```{r}
#| message: false
#| warning: false

# Begin to summarize the sorted data to see power consumption trends

# First, sort by customer type and category to look for grouped trends

summary_by_group <- power_filtered %>% 
  group_by(customer_type, category)

# Next, view the statistics for average kWh per day and wrap these into a 
# usable table format 

stat_power_table <- summary_by_group %>% 
  summarise(
    Total_Observations = n(),
    Mean_Daily_kWh = mean(kWh_per_day, na.rm = TRUE),
    SD_Daily_kWh = sd(kWh_per_day, na.rm = TRUE),
    Min_Daily_kWh = min(kWh_per_day, na.rm = TRUE),
    Max_Daily_kWh = max(kWh_per_day, na.rm = TRUE)
    )

stat_power_table %>% 
  kable(
    digits = 2,
    col.names = c("Type", "Category", "Mean Daily kWh", "SD Daily kWh", "Min Daily kWh", "Max Daily kWh"),
    caption = "Summary of Daily kWh Consumption by Type and Category"
  )

```

This table shows the general spread for each type and category of clients. Let's make a box plot to see this graphically and compare spread of data. 
Since the scales for the data are so different between types, show the 4 different box plots on different scales. Use this to examine the different categories in each type.

```{r}
#| message: false
#| warning: false
#| fig-alt: "A boxplot showing the distribution of daily electricity consumption in kWh for residential customers, grouped by connection type. Standard rates show a tighter distribution around 1 kWh, there is a concentration of high-use households around 4-5kWh that may be operating small businesses out of their residence"

# Figure 1: Residential
summary_by_group %>% 
  filter(customer_type == "Residential") %>% 
  ggplot(aes(x = category, y = kWh_per_day, fill = category)) + 
  geom_boxplot(outlier.alpha = 0.05, show.legend = FALSE) + 
  coord_cartesian(ylim = c(0, 7)) + 
  labs(title = "Residential Profiles", x = "Type", y = "Daily kWh") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1) 
  )


```
```{r}
#| message: false
#| warning: false
#| fig-alt: "A boxplot showing the distribution of daily electricity consumption in kWh for SME customers, grouped by type. Standard rates for most categories are below 10kWh, while several categories such as Mill and Factory are higher."


# Figure 2: SME
summary_by_group %>% 
  filter(customer_type == "SME") %>% 
  ggplot(aes(x = category, y = kWh_per_day, fill = category)) + 
  geom_boxplot(outlier.alpha = 0.05, show.legend = FALSE) + 
  coord_cartesian(ylim = c(0, 55)) + 
  labs(title = "Small Medium Enterprises", x = "Type", y = "Daily kWh") +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1) 
  )
```

```{r}
#| message: false
#| warning: false
#| fig-alt: "A boxplot showing the distribution of daily electricity consumption in kWh for Social Institutions , grouped by type. Standard rates for most categories are below 5kWh, with the largest usage from NGO and Carpentry."


# Figure 3: Social Institution
summary_by_group %>% 
  filter(customer_type == "Social Institution") %>% 
  ggplot(aes(x = category, y = kWh_per_day, fill = category)) + 
  geom_boxplot(outlier.alpha = 0.05, show.legend = FALSE) + 
  coord_cartesian(ylim = c(0, 15)) + 
  labs(title = "Social Institutions", x = "Type", y = "Daily kWh") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1) 
  )

```

```{r}
#| message: false
#| warning: false
#| fig-alt: "A boxplot showing the distribution of daily electricity consumption in kWh for Social Institutions , grouped by type. Standard rates for most categories are below 5kWh, with the largest usage from NGO and Carpentry."


# Figure 4: C&I
summary_by_group %>% 
  filter(customer_type == "C&I") %>% 
  ggplot(aes(x = category, y = kWh_per_day, fill = category)) + 
  geom_boxplot(outlier.alpha = 0.05, show.legend = FALSE) + 
  coord_cartesian(ylim = c(0, 500)) + 
  labs(title = "C&I Profiles", x = "Type", y = "Daily kWh") +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1) 
  )



```

Now that the spread of daily kWh has been assessed for each type and category, let's look at the general composition of daily power consumption. 

```{r}
#| message: false
#| warning: false
#| fig.width: 8
#| fig.height: 6
#| fig-alt: "A pie chart showing the overall composition of total electricity consumption across four customer types: Residential, SME, Social Institution, and C&I. Each slice is labeled with its corresponding percentage, clearly highlighting that SME accounts for the largest share of energy use, followed by Residential and C&I."

library(scales) 

# Summarize the data by customer_type
pie_data <- summary_by_group %>%
  group_by(customer_type) %>%
  summarise(total_kWh = sum(kWh_per_day, na.rm = TRUE)) %>%
  mutate(
    percentage = total_kWh / sum(total_kWh),
    label_pos = cumsum(total_kWh) - (total_kWh / 2) 
  )

# Build a stacked bar chart then a pie chart
ggplot(pie_data, aes(x = "", y = total_kWh, fill = customer_type)) +
  geom_bar(stat = "identity", width = 1, color = "white")+  
  coord_polar(theta = "y") + 
  geom_text(
    aes(y = label_pos, label = percent(percentage, accuracy = 0.1)), 
    size = 4, 
    color = "black"
  ) +
  
  # Clean up the background grid and axis lines from standard ggplot
  theme_void() + 
  labs(
    title = "Overall Composition of Daily kWh Consumption",
    subtitle = "Proportion of total energy usage across customer types",
    fill = "Customer Type"
  ) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
    plot.subtitle = element_text(hjust = 0.5, color = "gray40")
  )

```

It's also important to understand the peak kWh values to right-size the grid. This comes from the max kWh/h values, taking the maximum load during the highest use point during the day. Let's first group by day, then sum the value per day and see the composition from different types. 

# Conclusions

## Summary of Findings

## Questions and Next Steps

With a huge proportion of SMEs still uncategorized (NA), it is difficult to get accurate power consumption values from this which would be beneficial to scale based on the number of customers. Since there are some SME categories that differ significantly in power consumption, it is important to have as much granularity as possilbe to set predicted consumption values.

It would be good to understand the seasonal variation in power consumption, which will require sourcing more data over time. In addition, it would be interesting to review this data every year to see if average power consumption for residences and SMEs is increasing based on more reliable and affordable access to energy (such as buying more appliances etc.)

Finally, all of this data is from one site in DRC. Ideally, many sites would be included for comparison of regional consumption and see trends in types of customers. 

