---
title: "Solar mini-grid power consumption profiles in DRC"
author:
  - name: "Elyse McMillen"
    orcid: "0009-0002-2827-3062"
    email: "elyse.mcmillen@colorado.edu"
    affiliation:
      - name: "University of Colorado Boulder"
        department: "Department of Civil, Environmental and Architectural Engineering"
        city: "Boulder"
        state: "CO"
        country: "USA"
date: "July 02 2026"
format: html
editor: visual
---

# Introduction

This project explores existing power consumption data over a set range of dates from April - May 2026. The data was gathered from an operational solar mini-grid in the DRC providing power to a local community. The goal of the project is to summarize the consumption by category to be able to scale the expected consumption for other sites to predict mini-grid sizing in conflict-ridden areas with minimal on-the-ground data collection.

# Methods

After receiving the initial data from the IT team that was received from smart meters, the raw data was examined and processed.

## Reading the Data
```{r}
# Install packages
install.packages("readxl")
install.packages("writexl")
```

Since the file is a .xlsx file and not a .csv file, I had to install the "readxl" package. I had some initial challenges loading the "readxl" package but finally was able to get it to work after uninstalling and re-installing.

```{r}
# Load packages
library(tidyverse)
library(readxl)
library(dplyr)
library (writexl)

# Read data
raw_power_data <- read_excel(here::here("data/raw/Power_consumption_categorized_April_May_2026.xlsx"))

```

## Data Exploration Approach

Time to view the data and see what needs tidying!

```{r}
glimpse(raw_power_data)

dim(raw_power_data)

# calculate average daily kWh per customer type

customer_grouped <- raw_power_data %>% 
  group_by(`Customer type`) %>% 
  summarize(mean_daily_kWh = mean(`Daily kWh`, total_count = n()))


```

The initial data set is 165,588 rows and 7 columns. 
The column names are "date", "category", "customer code", "customer type", "hourly kWh average", "hourly kWh Max", and "daily kWh". I see some initial values showing of 0.000000 under some entries, and many repeating categories or types. 

Examined the average daily kWh for each major category for reference.

## Initial Data Tidying

Fortunately the dates are already in YYYY-MM-DD format, so R has been able to ingest those ok. 

Some of these are in French! 

I need to sort out any 0.000000 and NA entries to not skew the mean values for the categories. I also need to rename all of the column names that have spaces so I don't have to keep including "`" in every call-out.

```{r}

clean_power_data <- raw_power_data %>% 
  rename(
    customer_type = `Customer type`,
    category = Category,
    customer_code = `Customer Code`,
    customer_type = `Customer type`,
    kWh_avg_hourly = `Hourly kWh Average`,
    kWh_avg_max = `Hourly kWh Max`,
    kWh_per_day = `Daily kWh`
  )

power_na_clean <- clean_power_data %>% 
  mutate(across(where(is.character), ~ na_if(., "NA")))    # Fix text "NA"

na_counts <- power_na_clean %>%
  group_by(customer_type) %>%
  summarise(na_count = sum(is.na(category)))

```

The first time I tried to calculate the NA values I got 0... which is incorrect because I noted a bunch of NA values in the Category column. I first had to convert the text "NA" entries into actual NA entries. na_counts shows that there are 915 social institutions that are labeled NA. 

```{r}
# save as a .xlsx file 
write_xlsx(clean_power_data, here::here("data/processed/clean_power_data.xlsx"))

```


# Results

[This will be the core of your analysis with specific requirements]

# Conclusions

## Summary of Findings

## Questions and Next Steps
